---
title: "covid19-final-project"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)
library(lubridate)


```
## COVID-19 Overview
This project uses the New York Times COVID-19 data and imports the global and U.S. COVID-19 data. Data exploration and analysis is performed on the U.S. COVID-19 data.


``` {r data_import}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv")
urls <- str_c(url_in,file_names)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read_csv(urls[3])
US_deaths <- read_csv(urls[4])

```
## COVID-19 Tidying and Transformation

The following activities are performed to tidy and transform the data:

* Lat and Long columns are not required and are removed
* each date on a seperate row so that we have the cases per each date
* dates that have zero cases are removed
* cases and deaths are combined into one dataset

```{r data_tidy}

global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State','Country/Region', Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat,Long))




global_cases <- global_cases %>% filter(cases>0)

global <- global_cases %>% 
  full_join(global_deaths) %>%
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))
US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()


```
## COVID-19 Data Exploration and Visualization
Plotting the total number of cases in the U.S. along side the total number of deaths would indicate that the percentages of cases that are deaths remains fairly constant. 


```{r data_visual_us_cases_deaths}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()
           
US_totals %>% 
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y=deaths, color="deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)

state <- "New York"
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```
The three states with the highest number of deaths are:

* Mississippi
* Arizona
* Oklahoma

The three states with the lowest number of deaths are :
* American Samoa
* Northern Mariana Island
* Hawaii

```{r data_visual_deaths_by_state}


US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

par(mar=c(12,4,4,4))
# best 10 states
best_ten <- US_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
barplot(best_ten$deaths_per_thou, names.arg=best_ten$Province_State, ylab="Thousands of Deaths", main="10 States with the lowest death totals", las=3) 
  

# worst 10 states
worst_ten = US_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())


barplot(worst_ten$deaths_per_thou,names.arg=worst_ten$Province_State, ylab="Thousands of Deaths", main="10 States with the highest death totals", las=3) 
```  
## COVID-19 Data Modeling
At the U.S. level, the first visualization graph plotting both cases and deaths would indicate that total number of cases could help predict total number of deaths, it is shown in this section that at the state level this is connect be shown (at not with the linear regression model that is used). The Adjusted R squared is only 0.278 which would indicate that at the state level, the number of cases is a poor predictor of deaths and the graphs plotted the actually number of deaths and the predicted deaths show that the linear lined plotted by the predictions does not fit the actual deaths very well. This would indicate that there are significant other reasons as to why the number of deaths vary significantly between states. This other factors could be number of hospitals, access to resperators, or government public health policies.

```{r data_modeling}

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)

x_grid <- seq(1,380)
new_df <- tibble(cases_per_thou = x_grid)
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred %>% ggplot() +
  geom_point(aes(x= cases_per_thou, y = deaths_per_thou), color="blue") +
  geom_point(aes(x= cases_per_thou, y = pred), color="red")  
```