---
title: "NYPD Shooting Incident Data Report"
date: "2/5/2022"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
library(tidyverse)
library(lubridate)
```

## New York City (NYC) Shooting Incidents Are On The Rise
(Note: echo = TRUE so that R code may be seen as per course instructions)

Shooting incidents in NYC have generally been falling since 2011, however 2020 has shown a sharp increase in shooting incidents across the Brooklyn, the Bronx, Manhattan, Queens, and Staten Island with the highest increase being in Brooklyn. This is shown in the following diagram:

```{r shooting}

shooting <- read_csv("https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD")

shooting <- mutate(shooting, YEAR = year(mdy(OCCUR_DATE)), MONTH = month(mdy(OCCUR_DATE)), DAY = day(mdy(OCCUR_DATE)), MONTH_YEAR = str_c(month(mdy(OCCUR_DATE)), '-', year(mdy(OCCUR_DATE))))


total <- distinct(shooting %>% add_count(BORO, YEAR) %>% select(BORO,YEAR, n) %>% rename(SHOOTINGS = n))
xlabels = min(shooting$YEAR):max(shooting$YEAR)
```

```{r plot_shootings_by_year_boro}

ggplot(total, aes(x = YEAR, y = SHOOTINGS)) + 
  geom_line(aes(color = BORO, linetype = BORO)) +
  scale_x_continuous(labels=xlabels,breaks = xlabels) +
  theme(legend.position = "bottom") +
  labs(title = "Shootings by Year",
       x = "Year",
       y = "Shootings")
```
```{r plot_murders_by_year_boro}
murders <- filter(shooting, STATISTICAL_MURDER_FLAG == 'TRUE')
total_murders <- distinct(murders %>% add_count(BORO, YEAR) %>% select(BORO,YEAR, n) %>% rename(SHOOTINGS = n))
xlabels = min(murders$YEAR):max(murders$YEAR)
ggplot(total_murders, aes(x = YEAR, y = SHOOTINGS)) + 
  geom_line(aes(color = BORO, linetype = BORO)) +
  scale_x_continuous(labels=xlabels,breaks = xlabels) +
  theme(legend.position = "bottom") +
  labs(title = "Murders by Year",
       x = "Year",
       y = "Murders")
```
```{r model}
total <- distinct(shooting %>% add_count(BORO, YEAR) %>% select(BORO,YEAR, n) %>% rename(SHOOTINGS = n))
total_murders <- distinct(murders %>% add_count(BORO, YEAR) %>% select(BORO,YEAR, n) %>% rename(MURDERS = n))
tot = distinct(inner_join(total,total_murders))
tot <- mutate(tot, PERCENT_MUR=MURDERS/SHOOTINGS)
mod <- lm(PERCENT_MUR ~ SHOOTINGS, data = tot)
tot_w_pred <- tot %>% mutate(pred = predict(mod))

tot_w_pred %>% ggplot() +
  geom_point(aes(x = SHOOTINGS, y=PERCENT_MUR), color = "blue") +
  geom_point(aes(x = SHOOTINGS, y=pred), color = "red") +
  labs(title = "Graph: Model",
       x = "Year",
       y = "Murder Percentage")

```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r sessioninfo, echo=FALSE}
sessionInfo()
```